ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-02-02 paper.tex
Review of the last part

Given observed matrix data $\left\{T_j\right\}_{j\in J}$ (with a total
of $|J|$ observed matrix instances), for each observation $T_j$ the
conditional expectation at step $k$ is
%
\begin{equation}
Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_{++}^p}%
f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS%
,\label{eq:expdef}
\end{equation}
where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$,
$\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and
$f(T,S;\theta)$ is the full likelihood of the distribution with the
hidden scale parameter $S$.

We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$. Eliminating a global factor $\frac{1}{2}$ and constant terms which do not depend on the parameters in $\theta$ we have
\begin{equation}
\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]\\
-\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
\end{split}\label{eq:logf}
\end{equation}
The expectation computation requires computing the following two integrals:
\begin{equation}
\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k) S^{-1}\,dS.
\end{equation}
It can be done by using equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively, and this gives
\begin{equation}
\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
\end{equation}
\begin{equation}
\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
\end{equation}

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|.
\end{split}
\label{eq:fsimp}
\end{equation}
Here the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
\begin{equation}
W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

As $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$ is also SPD. The computation of $W_j^{(k)}$ can be done efficiently using Cholesky factorization via LAPACK routines and the BLAS routines for symmetric matrix update. When $p$ is much greater than $m$, i.e., the dimension of $\Sigma$ is much larger than $\Omega$, one can apply Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to reduce the cost of inverting the $p\times p$ matrix for each $W_j^{(k)}$. By using the formulae $W_j^{(k)}$ can be rewritten in the following equivalent form
\begin{equation}
W_j^{(k)}=\Sigma_k^{-1}-\Sigma_k^{-1}U_j\left(I_m+U_j^{\textrm{t}}\Sigma_k^{-1}U_j\right)^{-1}U_j^{\textrm{t}}\Sigma_k^{-1}\label{eq:SMW}
\end{equation}
with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper Cholesky factor of $\Omega_k$. Note that in Eq.~\eqref{eq:SMW} one inverts a $m\times m$ matrix for each $j$ and the inversion (or factorization) of $\Sigma_k$ only needs to be done once in each iteration.

@@

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.\label{eq:M}
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).\label{eq:Omega}
\end{equation}
For the rest of the parameter(s), we consider two possible situations.
\begin{itemize}
\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
\begin{equation}
\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}.\label{eq:Sigma}
\end{equation}
$\Sigma_{k+1}$ depends linearly on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
is a concave function of $\nu$ for all $\nu>0$. The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive. In this situation we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$ which has the form
\begin{equation}
h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log |\bar{W}^{(k)}|=0\label{eq:nu1}
\end{equation}
with $h(x):=p\log(x)-\psi_p(x/2)$.
\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$. The gradient equation of $\nu$ is given by
\begin{equation}
\log|\Sigma|-\log |\bar{W}^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
\end{equation}
\end{itemize}

When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood of matrix--$t$ distribution monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge due to the non-identifiability of matrix--$t$ mentioned at the end of section \ref{sec:matrixt}. This issue has also been discussed in \citep{2013arXiv1309.6609G} for ML estimation of matrix normal distributions. However, instead of fixing a single component in one of the scale matrices, we suggest setting the mean trace ratio
\begin{equation}
\frac{\textrm{tr}[\Omega_k]/m}{\textrm{tr}[\Sigma_k]/p}
\end{equation}
of $\Omega_k$ and $\Sigma_k$ to 1 by proper scaling at the end of each M-step. This ensures the convergence of the algorithm and also avoids possible numerical issues due to the possible large scale separation between $\Omega_k$ and $\Sigma_k$ during the iterations.

@@

For EM method, the initial values for the parameters has significant effect on the convergence behavior as the algorithm does not guarantee reaching the global optimum. While this is less of an issue for matrix--$t$ distributions as the ML estimation is a convex optimization problem in parameters $\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, it is still desirable to have an efficient method for selecting good initial values to reduce the number of EM iterations for reaching the required accuracy as the classical log EM is a first order algorithm.

For the location parameter $M$, the component-wise medians and spatial median \citep{Vardi15022000} with distance induced by Frobenius norm
\begin{equation}
d(T,T')=\sqrt{\textrm{tr}\left[(T-T')(T-T')^{\textrm{t}}\right]}
\end{equation}
are proper choices we found in the implementation. Both estimators can be computed with time complexity $O(Nmp)$ with $N$ the number of matrices in the given data.

For the rest of the parameters, we center the data by the initial values of $M$ and use the property of low dimensional projection of matrix--$t$ distributions to obtain the initial values. It is known that if $T\sim T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, then $Tv$ is multivariate $t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega v}{\nu}\right)\Sigma$ and the same degress of freedom parameter $\nu$ for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}. This allows using efficient routines for estimating multivariate $t$ distributions to retrieve the initial values.

@@

\section*{Acknowledgement}
This work is not supported by any grant and has not been presented in any regional or international meetings.

>>>>

Given observed matrix data $\left\{T_j\right\}_{j\in J}$ (with a total
of $|J|$ observed matrix instances), for each observation $T_j$ the
conditional expectation at step $k$ is
%
\begin{equation}
Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_{++}^p}%
f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS%
,\label{eq:expdef}
\end{equation}
where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$,
$\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and
$f(T,S;\theta)$ is the full joint likelihood of the distribution with
the hidden scale parameter $S$.

We can split the integral in Eq.~\eqref{eq:expdef} by the terms of
$\log f(T_{j},S;\theta)$. Eliminating a global factor $\frac{1}{2}$
and constant terms which do not depend on the parameters in $\theta$
we have
\begin{equation}
\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]\\
-\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
\end{split}\label{eq:logf}
\end{equation}
The first two terms in Eq.~\ref{eq:logf} require computing the
following integrals:
\begin{equation}
\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad
\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k) S^{-1}\,dS.
\end{equation}
Note that the trace operation commutes with the matrix integration in
the for the second term.  The integrals can be computed from
identities obtained by taking the derivative of
Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix
$A$, respectively:
%
\begin{equation}
\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
\end{equation}
\begin{equation}
\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
\end{equation}
For the rest of the terms in Eq.~\eqref{eq:logf}, the integration over
$S$ simply cancels out the prefactor.

Thus, using the equalities Eq.~\eqref{eq:intderivative1} and
Eq.~\eqref{eq:intderivative2}, and after some rearrangements, we
obtain the closed form expression of $Q_j(\theta,\theta_k)$, where we
have dropped the irrelevant constant terms
%
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|.
\end{split}
\label{eq:fsimp}
\end{equation}
Here the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
\begin{equation}
W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}
Since $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$
is also SPD.

The main computational challenge in our algorithm is calculating the
$W_j^{(k)}$ matrices efficiently. We employ Cholesky factorization via
LAPACK and the BLAS routines for symmetric
matrix update. When $p$ is much greater than $m$, i.e., the dimension
of $\Sigma$ is much larger than $\Omega$, one can apply the
Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to
reduce the cost of inverting the $p\times p$ matrix for each
$W_j^{(k)}$. By using the formulae $W_j^{(k)}$ can be written as
%
\begin{equation}
W_j^{(k)}=\Sigma_k^{-1}-\Sigma_k^{-1}U_j\left(I_m+U_j^{\textrm{t}}\Sigma_k^{-1}U_j\right)^{-1}U_j^{\textrm{t}}\Sigma_k^{-1}\label{eq:SMW}
\end{equation}
with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper
Cholesky factor of $\Omega_k$. Note that in Eq.~\eqref{eq:SMW} one
inverts a $m\times m$ matrix for each $j$ and the inversion (or
factorization) of $\Sigma_k$ only needs to be done once in each
iteration.

@@

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute
the gradient of each parameter and equate them to $0$ to obtain
$\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the
input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.\label{eq:M}
\end{equation}
Upon maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$, we obtain
$\Omega_{k+1}$
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).\label{eq:Omega}
\end{equation}
For the rest of the parameter(s), we consider two possible situations.
\begin{itemize}
\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing
	$Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain
	$\Sigma_{k+1}$ in terms of the inverse of the mean of
	$W_j^{{(k)}}$
\begin{equation}
\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}.\label{eq:Sigma}
\end{equation}
$\Sigma_{k+1}$ depends linearly on $\nu$. Due to this, we substitute
$\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with
respect to $\nu$. We note that
\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
is a concave function of $\nu$ for all $\nu>0$. In this
situation we obtain $\nu_{k+1}$ by solving the equation
$\tilde{Q}'(\nu)=0$ which has the form
\begin{equation}
h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log |\bar{W}^{(k)}|=0\label{eq:nu1}
\end{equation}
with $h(x):=p\log(x)-\psi_p(x/2)$.
\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$. The gradient equation of $\nu$ is given by
\begin{equation}
\log|\Sigma|-\log |\bar{W}^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
\end{equation}
\end{itemize}

When $\Omega$ and $\Sigma$ are both unknown, the updating steps given
in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood
of matrix--$t$ distribution monotonically but the parameters
$\Omega_k$ and $\Sigma_k$ may not converge due to the
non-identifiability of matrix--$t$ mentioned at the end of section
\ref{sec:matrixt}. This issue has also been discussed in
\citep{2013arXiv1309.6609G} for ML estimation of matrix normal
distributions. However, instead of fixing a single component in one of
the scale matrices, we suggest setting the mean trace ratio
\begin{equation}
\frac{\textrm{tr}[\Omega_k]/m}{\textrm{tr}[\Sigma_k]/p}
\end{equation}
of $\Omega_k$ and $\Sigma_k$ to 1 by proper scaling at the end of each
M-step. This ensures the convergence of the algorithm and also avoids
possible numerical issues due to the possible large scale separation
between $\Omega_k$ and $\Sigma_k$ in the iterative steps.

@@

In the EM method, the initial values of the parameters have
a significant effect on the convergence behavior, since the algorithm
does not guarantee reaching the global optimum. This is less of an
issue for matrix--$t$ distributions because the ML estimation is
a convex optimization problem in the parameters
$\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, but it is still
desirable to select good initial values to reduce the number of EM
iterations necessary to reach the required accuracy as the classical
log EM is a first order algorithm.

For the location parameter $M$, we found that the component-wise
median, and spatial
median \citep{Vardi15022000} with distance induced by Frobenius norm
\begin{equation}
d(T,T')=\sqrt{\textrm{tr}\left[(T-T')(T-T')^{\textrm{t}}\right]}
\end{equation}
are good choices. Both estimators can be computed with time complexity
$O(Nmp)$ with $N=|J|$ the number of matrices in the given data.

To obtain the rest of initial the parameters, we first center the data
by the initial values of $M$. We then  use the property of low
dimensional projection of matrix--$t$ distributions to obtain the
initial values.  It is known that if $T\sim
T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, then $Tv$ is a multivariate
$t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega
v}{\nu}\right)\Sigma$ and the same degrees of freedom parameter $\nu$
for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
This allows using existing routines for estimating multivariate $t$
distributions to retrieve the initial values.

@@

\section*{Acknowledgement}
Z.K. is supported by  the  Executive  Research  Agency  (REA)  of  the
European Commission under the Grant Agreement PITN-GA-2012-316704
(HiggsTools).
